---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/README-"
)
```

# galah

[![Travis-CI Build Status](https://travis-ci.com/AtlasOfLivingAustralia/galah.svg?branch=master)](https://travis-ci.com/AtlasOfLivingAustralia/galah)
[![codecov](https://codecov.io/gh/AtlasOfLivingAustralia/galah/branch/master/graph/badge.svg)](https://codecov.io/gh/AtlasOfLivingAustralia/galah)

`galah` is an R package for accessing data from the [Atlas of Living Australia (ALA)](https://ala.org.au) [APIs](https://api.ala.org.au/). It is the successor to [`ALA4R`](https://github.com/AtlasOfLivingAustralia/ALA4R)

The ALA aggregates Australian biodiversity data from a range of sources, and makes the data freely available for scientists, policy makers, industry and the general public.

If you have any questions or suggestions, please [contact us](mailto:support@ala.org.au).

## Citing the package
Please cite `galah`. To generate a citation for the package version you are using, you can run
```r
citation(package = "galah")
```

## Citing data
If you're using occurrence data downloaded through `galah` in a publication, please generate a DOI and cite it. 
To generate a DOI, set `mint_doi = TRUE` in a call to `ala_occurrences`.
To generate a citation using the dataset, use `ala_citation`

```{r eval = FALSE}
occ <- ala_occurrences(..., mint_doi = TRUE)
doi <- attr(occ, "doi")
cit <- ala_citation(occ)
```

## Installing

`galah` is not yet available on CRAN. To install the package from GitHub:

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("AtlasOfLivingAustralia/galah")
```

On Linux you will first need to ensure that `libcurl` and `v8` (version <= 3.15) are installed on your system --- e.g. on Ubuntu/Debian, open a terminal and do:

```{sh eval=FALSE}
sudo apt-get install libcurl4-openssl-dev libv8-3.14-dev
```

## Getting started

- The `galah` [vignette](https://atlasoflivingaustralia.github.io/galah/articles/galah.html) provides an overview of the main features
- Run `?galah` or view the [reference page](https://atlasoflivingaustralia.github.io/galah/reference/index.html) to view all functions.

