---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/README-"
)
```

# koala

[![Travis-CI Build Status](https://travis-ci.com/AtlasOfLivingAustralia/koala.svg?branch=master)](https://travis-ci.com/AtlasOfLivingAustralia/koala)
[![AppVeyor Build status](https://ci.appveyor.com/api/projects/status/g9pudc4l7053w4vn/branch/master?svg=true)](https://ci.appveyor.com/project/PeggyNewman/ala4r/branch/master)
[![codecov](https://codecov.io/gh/AtlasOfLivingAustralia/koala/branch/master/graph/badge.svg)](https://codecov.io/gh/AtlasOfLivingAustralia/koala)

<code>koala</code> enables the R community to directly access data hosted by the Atlas of Living Australia (ALA).

The ALA aggregates Australian biodiversity data from a range of sources, and makes the data freely available for scientists, policy makers, industry and the general public.

If you have any questions or suggestions, please [contact us](mailto:support@ala.org.au).

## Citing koala
To generate a citation for koala run:
```r
citation(package = "koala")
```

## Installing

To install the development version from GitHub:

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("AtlasOfLivingAustralia/koala")
```

On Linux you will first need to ensure that `libcurl` and `v8` (version <= 3.15) are installed on your system --- e.g. on Ubuntu/Debian, open a terminal and do:

```{sh eval=FALSE}
sudo apt-get install libcurl4-openssl-dev libv8-3.14-dev
```


## Usage

See the online documentation at https://atlasoflivingaustralia.github.io/koala/ including the [package vignette](https://atlasoflivingaustralia.github.io/koala/articles/koala.html).
