---
title: "galah"
author: "Matilda Stevenson"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{galah}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
options(width=120)
```

# About
`galah` is an R interface to biodiversity data hosted by the [Atlas of Living Australia](https://ala.org.au) (ALA). The ALA is a repository of biodiversity
data, focussed primarily on observations of individual life forms. Like the
Global Biodiversity Information Facility ([GBIF](https://www.gbif.org)), the
basic unit of data at ALA is an **occurrence** record, based on the ['Darwin Core'](https://dwc.tdwg.org) data standard.

`galah` enables users to locate and download species observations, taxonomic
information, or associated media such images or sounds, and to restrict their
queries to particular taxa or locations. Users can specify which columns are
returned by a query, or restrict their results to observations that meet
particular quality-control criteria. All functions return a `data.frame` as
their standard format.

Functions in `galah` are designed according to a nested architecture. Users
that require data should begin by locating the relevant `ala_` function (see
[downloading data section](#downloading-data)); the arguments within that
function then call correspondingly-named [`select_` functions](#filtering-data);
and finally the specific values that can be interpreted by those `select_`
functions are given by `find_` functions.

## Installation

Install from CRAN:
```{r eval=FALSE}
install.packages("galah")
```

Install the development version from GitHub:
```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("AtlasOfLivingAustralia/galah")
```

See the [README](https://github.com/AtlasOfLivingAustralia/galah/blob/master/README.md) for system requirements.

Load the package
```{r}
library(galah)
```

# Filtering data

Each occurrence record contains taxonomic information, and also some
information about the observation itself, such as its location and the date
of the observation. Each piece of information associated with a
given occurrence is stored in a **field**,  which corresponds to a **column**
when imported to a `data.frame`.

Data fields are important because they provide a means to *filter*
occurrence records;  i.e. to return only the information that you need, and
no more. Consequently, much of the architecture of `galah` has been
designed to make filtering as simple as possible, by using functions with the
`select_` prefix.

## Taxonomic filtering
`select_taxa()` enables users search for taxonomic names and check the results
are 'correct' before using the result to download data.
The function allows both free-text searches and searches where the rank(s) are
specified. Specifying the rank can be useful when names are ambiguous.
```{r}
# free text search
taxa_filter <- select_taxa("Eolophus")

# specifying ranks
select_taxa(query = list(genus = "Eolophus", kingdom = "Aves"))
```
`select_taxa()` can optionally provide information about child concepts, and
counts of the number of records held by the ALA for the specified taxa.
```{r}
select_taxa(query = "Eolophus", children = TRUE, counts = TRUE)
```
This shows that there is only one species in the family *Eolophus*.

## Location-based filtering
Users can provide an `sf` object or a Well-Known Text (WKT) string for
location-based filtering.
```{r eval = FALSE}
locations <- select_locations(query = st_read('act_rect.shp'))
```

## Field based filtering
As mentioned above, all occurrence records in the ALA contain additional
information about the record, stored in **fields**. Field-based filters are
specified with `select_filters()`, which takes indvidual filters, in the form
`field = value`, and/or a [data quality profile](#data-quality-profiles).

To find available fields and corresponding valid values, field lookup 
functions are provided. For finding field names, use `search_fields()`, for
finding valid field values, use `find_field_values()`.
```{r}
search_fields("basis")
field_values <- find_field_values("basisOfRecord")
```
Build a field filter

```{r eval = FALSE}
filters <- select_filters(basisOfRecord = "HumanObservation")
```

By default, a filter is included. To negate a filter, use `exclude()`.
```{r}
filters <- select_filters(basisOfRecord = "HumanObservation",
                          occurrenceStatus = exclude("absent"))
```


### Data quality profiles
A notable extention of the filtering approach is to remove records with low
'quality'. ALA performs quality control checks on all records that it stores.
These checks are used to generate new fields, that can then be used to filter
out records that are unsuitable for particular applications. However, there
are many possible data quality checks, and it is not always clear which are
most appropriate in a given instance. Therefore, `galah` supports ALA
data quality **profiles**, which can be passed to `select_filters()`to quickly
remove undesirable records. A full list of data quality profiles is returned by
`find_profiles()`.

```{r}
profiles <- find_profiles()
```
View filters included in a profile
```{r}
find_profile_attributes("ALA")
```
Include a profile in the filters
```{r}
filters <- select_filters(basisOfRecord = "HumanObservation",
                          profile = "ALA")
```

# Downloading data

Functions that return data from ALA are named with the prefix `ala_`,
followed by a suffix describing the information that they provide.

By combining different filter functions, it is possible to build complex
queries to return only the most valuable information for a given problem.
Once you have retrieved taxon information, you can use this to [search for
occurrence records](#occurrence-data) with `ala_occurrences()`. However, it is
also possible to [download data on species](#species-data) via `ala_species()`,
or [media content](#media_downloads) (largely images) via `ala_media()`.
Alternatively, users can [retrieve record counts](#record-counts) using `ala_counts()`.

## Occurrence data

In addition to the [filter functions](#filtering-data) above, when downloading
occurrence data users can specify which columns are returned using
`select_columns()`. Individual column names and/or column groups can be
specified.
To view the fields for each group, see the documentation for `select_columns()`.
To view the list of available fields, run `search_fields()`.
```{r eval = FALSE}
cols <- select_columns("institutionID", group = "basic")
```

To download occurrence data you will need to specify your email in
`ala_config()`. This email must be associated with an active ALA account. See
more information in the [config section](#config)
```{r include = FALSE}
ala_config(email = "ala4r@ala.org.au")
```

```{r eval = FALSE}
ala_config(email = your_email_here, profile_path = path_to_profile)
```

```{r include = FALSE}
occ <- read.csv("eolophus_roseicapilla.csv")
```

Download occurrence records for *Eolophus roseicapilla*
```{r eval = FALSE}
occ <- ala_occurrences(taxa = select_taxa("Eolophus roseicapilla"),
                       filters = select_filters(stateProvince = "Australian Capital Territory",
                                                year = seq(2010, 2020),
                                                profile = "ALA"),
                       columns = select_columns("institutionID", group = "basic"))
```

```{r}
head(occ)
```

## Species data
A common use case of the ALA is to identify which species occur in a specified
region, time period, or taxonomic group. `ala_species()` enables the user to 
look up this information, using the common set of filter functions.
```{r include = FALSE}
ala_config(cache_directory = tempdir())
```

```{r}
# List rodent species in the NT
species <- ala_species(taxa = select_taxa("Rodentia"),
            filters = select_filters(stateProvince = "Northern Territory"))
head(species)
```

## Record counts
`ala_counts()` provides summary counts on records in the ALA, without needing
to download all the records. In addition to the filter arguments, it has an
optional `group_by` argument, which provides counts binned by the requested
field.
```{r warning = FALSE}
# Total number of records in the ALA
ala_counts()

# Total number of records, broken down by kindgom
ala_counts(group_by = "kingdom")
```


## Media downloads
In addition to text data describing individual occurrences and their attributes, ALA stores images, sounds and videos associated with a given record. These can be downloaded to `R` using `ala_media()` and the same
set of filters as the other data download functions.
```{r warning=FALSE, message=FALSE, eval = FALSE}
# Use the occurrences previously downloaded
media_data <- ala_media(
     taxa = select_taxa("Eolophus roseicapilla"),
     filters = select_filters(year = 2020),
     download_dir = "media")
```


# Config

Various aspects of the galah package can be customized. To preserve
configuration for future sessions, set `profile_path` to a location of a
`.Rprofile` file.

## Email
To download occurrence records, you will need to provide an email address
registered with the ALA. You can create an account [here](https://auth.ala.org.au/userdetails/registration/createAccount).
Once an email is registered with the ALA, it should be stored in the config:
```{r eval=FALSE}
ala_config(email="myemail@gmail.com")
```

## Caching
`galah` can cache most results to local files. This means that if the same code
is run multiple times, the second and subsequent iterations will be faster.

By default, this caching is session-based, meaning that the local files are
stored in a temporary directory that is automatically deleted when the R
session is ended. This behaviour can be altered so that caching is permanent,
by setting the caching directory to a non-temporary location.

```{r eval=FALSE}
ala_config(cache_directory="example/dir")
```

By default, caching is turned off. To turn caching on, run
```{r eval=FALSE}
ala_config(caching=FALSE)
```


## Debugging
If things aren't working as expected, more detail (particularly about web requests and caching behaviour) can be obtained by setting the `verbose` configuration option:

```{r eval=FALSE}
ala_config(verbose=TRUE)
```

## Setting the download reason
ALA requires that you provide a reason when downloading occurrence data (via the galah `ala_occurrences()` function). The reason is set as "scientific research" by default, but you can change this using `ala_config()`. See `find_reasons()` for valid download reasons.

```{r eval=FALSE}
ala_config(download_reason_id=your_reason_id)
```
