---
title: "International Atlases"
author: "Matilda Stevenson"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{International Atlases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The software developed by the ALA is now used by 27 International Living Atlases. 
Although there are differences in the versions of core components used by the
Atlases, and variations in the web services available, `galah` is flexible enough to 
handle most of these differences. Below we explain how to change the Living
Atlas the package points to, and how to put in a pull request if you would like
to see another Living Atlas supported in `galah`.

## Supported Atlases
To show the Atlases currently supported by `galah`, use `find_atlases()`. This
function also provides information on which taxonomic source each Atlas uses.
```{r}
library(galah)
find_atlases()
```

## Set Atlas
Use `ala_config()` to set the Atlas to use. This will automatically populate the 
server configuration for your selected Atlas. By default, the atlas is Australia.
If you intend to download records you will need to register with the relevant
Atlas. 

```{r eval=FALSE}
ala_config(atlas = "UK", email = "your_email_here")
```

## Get data for another atlas
```{r}
ala_counts()
```

## Using taxonomic filters for another atlas
`galah` only supports searching for Australian taxonomy; for other Atlases,
`taxize` should be used to search for the taxonomic id, and this id can be passed
to the `taxa` argument of `ala_` functions. For example:
```{r,error=TRUE}
library(taxize)
# Use UK taxonomy
id <- get_nbnid("Vulpes vulpes", rows = 1)
ala_counts(taxa = id)
```

## Retrieving occurrence records
```{r}
# Download records from the Austrian atlas
ala_config(atlas = "Austria")
id <- get_gbifid("bombina variegata", rows = 1)
ala_counts(taxa = id)
```

```{r include = FALSE}
occ <- read.csv('bombina_variegata.csv')
```

```{r eval = FALSE}
occ <- ala_occurrences(taxa = id)
```

```{r}
head(occ)
```

## Supporting another Atlas
`galah` currently supports a subset of the existing Living Atlases. If you would
like to see another Living Atlas supported, add the server configuration for the 
Atlas in the `R/find_atlases.R` file, and put in a pull request.
