---
title: "Taxonomic information"
author: "Matilda Stevenson"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taxonomic information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`galah` provides multiple ways of retrieving taxonomic information, and the best
method depends of the type of information required. Below, some use cases are
outlined. 

## Searching for taxonomic information

To search through the ALA's taxonomic backbone, use `search_taxa`. It takes a
string search term, and provides the higher ranks of the search term. 
```{r}
library(galah)
search_taxa("Mammalia")
```

`search_taxa` can also provide information on taxa below the provided search
term, enabling construction of a taxonomic tree. For example, we can build and
visualise all the families in class **Mammalia**. 
```{r}
library(collapsibleTree)
taxa <- search_taxa("Mammalia", downto = "family")
collapsibleTree(
  taxa,
  hierarchy = names(taxa)[-which(names(taxa) %in% c("kingdom", "phylum", "class"))],
  root = "Mammalia"
  )
```

Note that if you are traversing more than 3 ranks, it might be more efficient
to use `ala_species`, as `search_taxa` is efficient when the search tree
is large.

## Search for species matching criteria


## Species count information


## Taxonomic filtering

`ala_species` is for getting species matching a set of filters. If a user wants
a different rank it is fairly trivial to group the data themselves.
